---
title: "P8131_HW10"
author: "Pei Tian, pt2632"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F)
```

```{r lib-import}
library(tidyverse)
library(KMsurv)
library(survival)
library(survminer)
```

# Problem 1
```{r 1-data}
data1 = tibble(
  time = c(4, 12, 15, 21, 23, 2, 6, 8, 10, 19),
  cens = c(1, 0, 1, 0, 1, 1, 0, 0, 1, 1), 
  group = c(rep(1, 5), rep(2, 5)) |> factor()
)
ggsurvplot(survfit(Surv(time, cens) ~ group, data = data1), 
           data = data1, pval = TRUE, conf.int = TRUE,
           palette = c("tomato", "steelblue"),
           xlab = "Time", ylab = "Survival probability")
```

```{r 1-logrank}
survdiff(Surv(time, cens) ~ group, data = data1)
```

**Interpretation: **
Since the p-value is higher than 0.05, it means that we didn't find a strong enough difference between the two groups. So we can't reject the $H_0$ in 0.05 significant level. 


# Problem 2
```{r 2-survplot}
data("kidtran")

# male
ggsurvplot(survfit(Surv(time, delta) ~ race, data = subset(kidtran, gender == 1)), 
           pval = TRUE, conf.int = TRUE,
           palette = c("tomato", "steelblue"),
           xlab = "Time", ylab = "Survival probability", 
           legend.title = "Race",
           legend.labs = c("White", "Black"),
           title = "Male")

# female
ggsurvplot(survfit(Surv(time, delta) ~ race, data = subset(kidtran, gender == 2)), 
           pval = TRUE, conf.int = TRUE,
           palette = c("limegreen", "orange"),
           xlab = "Time", ylab = "Survival probability", 
           legend.title = "Race",
           legend.labs = c("White", "Black"),
           title = "Female")
```
```{r 2-logrank-male}
survdiff(Surv(time, delta) ~ race, data = subset(kidtran, gender == 1))
```


```{r 2-logrank-female}
survdiff(Surv(time, delta) ~ race, data = subset(kidtran, gender == 2))
```
**Interpretation: **

- For male group, the p-value is higher than 0.05, suggesting that there is no significant difference in survival times between the two races for the male group in 0.05 significance level. 

- For female group, the p-value is less than 0.05, suggesting that there is significant difference in survival times between the two races for the female group in 0.05 significance level.


# Problem 3
```{r 3-ph}
data("larynx")
larynx = larynx |> 
  mutate(z1 = (stage == 2) |> as.integer() |> factor(),
         z2 = (stage == 3) |> as.integer() |> factor(),
         z3 = (stage == 4) |> as.integer() |> factor())
ph_model = coxph(Surv(time, delta) ~ z1 + z2 + z3 + age + z1 : age, data = larynx, ties = "breslow")
beta = coef(ph_model)
ph_model
```

**Interpretation: **

- `Z1`: With fixed age and specific time point, the log hazard ratio will change `r beta[1]` for patient in stage II comparing to that in stage I, and the effect is statistically significant (p < 0.05).

- `Z2`: With fixed age and specific time point, the log hazard ratio will change `r beta[2]` for patient in stage III comparing to that in stage I, and the effect is not statistically significant (p > 0.05).

- `Z3`: With fixed age and specific time point, the log hazard ratio will change `r beta[3]` for patient in stage VI comparing to that in stage I, and the effect is statistically significant (p < 0.05).

- `age`: With specific time point, the log hazard ratio will change `r beta[4]` with one unit change in age, and the effect is not statistically significant (p > 0.05).

- `Z1:age`: With specific time point, the log hazard ratio of patient in stage II will change `r beta[5]`  with one unit change in age, and the effect is statistically significant (p < 0.05).


The overall model test (likelihood ratio test) is statistically significant (p < 0.001), indicating that the model as a whole is a significant improvement over the null model (a model with no predictors).

The model is:
\(h(t) = h_0(t) \exp(\beta_1 z_1 + \beta_2 z_2 + \beta_3 z_3 + \beta_4 \text{age} + \beta_5 z_1 \times \text{age})\)


The relative risk of dying for a stage II patient of age 50 as compared to a stage I patient of the same age is \(e^{\beta_1 + 50 * \beta_5}\)  = `r exp(beta[1] + 50 * beta[5])`.
