---
title: "P8131_HW5"
author: "Pei Tian, pt2632"
date: "`r Sys.Date()`"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F)
```

```{r lib-import, message=FALSE, include=FALSE}
library(tidyverse)
library(pscl)
```

## Problem 1
```{r data1}
data1 = read_table("./data/HW5-crab.txt") |>
  janitor::clean_names()
```
(a)
```{r model1-1}
m1 = glm(sa ~ w, family = poisson, data = data1)
summary(m1)
```
```{r inference1-1}
coef1 = coef(m1)
G1 = sum(residuals(m1, type = "pearson")^2)
df1 = m1$df.residual
pval = 1 - pchisq(G1, df1)
pval
```
**Interpretation: ** Denote the count of satellites as $Y$, and $Y\sim Pois(\lambda)$, so the fitted model is \(\log \lambda = \) `r coef1[1]` + `r coef1[2]` * w. The log count of satellites will increase `r coef1[2]` when w increases by 1 unit, holding other factors constant.


(b)
```{r model1-2}
m2 = glm(sa ~ w + wt, family = poisson, data = data1)
df2 = m2$df.residual
cmp_pval = 1 - pchisq(m1$deviance - m2$deviance, df1 - df2)
```
**Interpretation: ** With the deviance analysis, we could find that p-value is `r cmp_pval` < 0.05, so we should reject the null hypothesis and choose the larger model (M2). 

(c)
```{r over-dispersion}
resid = residuals(m2, type = "pearson")
phi = sum(resid^2) / m2$df.residual
phi
```

```{r half-normal-plot}
n = length(resid)
data1 |> 
  ggplot(aes(x = qnorm((n + 1:n + 0.5) / (2 * n + 1.125)))) +
  geom_point(aes(y = sort(abs(resid)))) +
  geom_abline(slope = 1, intercept = 0) +
  geom_abline(slope = sqrt(phi), intercept = 0, lty = 2) + 
  theme_bw() + 
  labs(
    title = "Half-Normal Plot",
    x = "Expected Half-Normal Order Stats", 
    y = "Ordered Abs Pearson Residuals"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  ) 
```


```{r overdispersion-adjust}
result = summary(m2, dispersion = phi)
coef2 = coef(result)
result
```
**Interpretation: **From the fitting result with over-dispersion adjustment, we could find that \(\log \lambda = \) `r coef2[1, 1]` + `r coef2[2, 1]` * w + `r coef2[3, 1]` * wt. The log count of satellites will increase `r coef2[2, 1]` when w increases by 1 unit, and that will increase `r coef2[3, 1]` when wt increase by 1 unit, holding other factors constant.

## Problem 2
```{r data2}
data2 = read.table("./data/HW5-parasite.txt", header = T) |>
  janitor::clean_names() |>
  mutate(year = factor(year),
         area = factor(area))
```

(a)
```{r model2}
model = glm(intensity ~ area + year + length, family = poisson, data = data2)
coef3 = coef(model)
summary(model)
```
**Interpretation: **Denote the count of parasites as $Y$, then $Y \sim Pois(\lambda)$. From the model fitting result, we could conclude that the log count of parasites will increase `r coef3[2]` when area is 2, `r coef3[3]` when area is 3 and `r coef3[4]` when area is 4 (reference category with area as 1), holding other factors constant. 

Meanwhile, the log count of parasites will increase `r coef3[5]` when the year is 2000, `r coef3[6]` when the year is 2001 (reference category with year as 1999), holding other factors constant. 

And the log count of parasites will increase `r coef3[7]` when length increase by 1 unit, holding other factors constant.


(b)
```{r goodness}
resid = residuals(model, type = "pearson")
G = sum(resid ^ 2)
pval = 1 - pchisq(G, df = model$df.residual)
```
The Pearson statistics of model is `r G`, and the p-value is `r pval` < 0.05, so we could conclude that fitting result is not good.

(c)
```{r}
inflated_model = zeroinfl(intensity ~ area + year + length | area + year + length, data = data2)
coef4 = coef(inflated_model)
summary(inflated_model)
```
**Interpretation: ** 

1. Poisson Model:

The log count of parasites is expected to increase by `r coef4[2]` for Area 2 compared to the baseline Area 1, holding other factors constant.
The log count of parasites is expected to increase by `r coef4[3]` for Area 3 compared to the baseline Area 1, holding other factors constant.
The log count of parasites is expected to increase by `r coef4[4]` for Area 4 compared to the baseline Area 1, holding other factors constant.

The log count of parasites is expected to increase by `r coef4[5]` in the year 2000 compared to the baseline year of 1999, holding other factors constant.
The log count of parasites is expected to increase by `r coef4[6]` in the year 2001 compared to the baseline year of 1999, holding other factors constant.

The log count of parasites is expected to decrease by `r coef4[7]` for each one-unit increase in fish length, holding other factors constant. 

2. Binomial Model: 

The log odds of having zero parasites for Area 2 compared to Area 1 is `r coef4[9]`, holding other factors constant.
The log odds of having zero parasites for Area 3 compared to Area 1 is `r coef4[10]`, holding other factors constant.
The log odds of having zero parasites for Area 4 compared to Area 1 is `r coef4[11]`, holding other factors constant.

The log odds of having zero parasites for the year 2000 compared to 1999 is `r coef4[12]`, holding other factors constant.
The log odds of having zero parasites for the year 2001 compared to 1999 is `r coef4[13]`, holding other factors constant.

The log odds of having zero parasites will increase `r coef4[14]` when length increase by 1 unit, holding other factors constant.


