---
title: "P8131_HW3"
author: "Pei Tian, pt2632"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r lib-import, message=FALSE, include=FALSE}
library(tidyverse)
```

## Problem 1
```{r data1}
case = c(0, 5, 21, 34, 36, 8, 1, 4, 25, 42, 19, 5)
control = c(106, 164, 138, 139, 88, 31, 9, 26, 29, 27, 18, 0)
data1 = tibble(
  age = rep(c(25, 35, 45, 55, 65, 75), 2),
  alcohol = c(rep("0-79 g", 6), rep("80+ g", 6)) |> factor(),
  resp = cbind(case, control)
)
```
(a)
```{r 1a}
model_a = glm(resp ~ age + alcohol, family = binomial(link = 'logit'), data = data1)
coef1 = summary(model_a)$coefficient

model_a |> 
  broom::tidy() |>
  knitr::kable(digits = 3)
```
**Interpretation:** denoting the probability of case occurring as $\pi$, then we have $\log \frac{\pi}{1-\pi}$ = `r coef1[1]` + `r coef1[2]` * age + `r coef1[3]` * alcohol_consumption. One unit change of age will lead to `r coef1[2]` units change of log odds ratio, while the people consuming alcohol >80g will have `r coef1[3]` units log odds ratio to become a patient.


(b) Denote the probability of esophageal cancer as $\pi$, then
$$
M_0: \log \frac{\pi}{1-\pi} = \beta_0 + \beta_1*age
$$

$$
M_1: \log \frac{\pi}{1-\pi} = \beta_0 + \beta_1*age + \beta_2*alcohol
$$

```{r 1b}
m0 = glm(resp ~ age, family = binomial(link = 'logit'), data = data1)
dev0 = sum(residuals(m0,type='deviance')^2)
dev1 = sum(residuals(model_a,type='deviance')^2)
n = data1$resp
pval = 1 - pchisq(dev0 - dev1, 1) 
```
After delivering deviance analysis on 2 model, the p-value=`r pval` is less than 0.01, so we have 99% confidence that 2 models are significant different. Therefore, $M_1$ has better performance than $M_0$, $M_1$ is more reasonable.

## Problem 2
```{r data2}
bean_75 = c(10, 23, 23, 26, 17)
cucumber_75 = c(5, 53, 55, 32, 46, 10)
bean_73 = c(8, 10, 8, 23, 0)
cucumber_73 = c(3, 22, 15, 32, 3)
batch_count = c(c(39, 62, 81, 51, 39), 
                c(6, 74, 72, 51, 79, 13),
                c(16, 30, 28, 45, 4),
                c(12, 41, 30, 51, 7))
data2 = tibble(
  seed = factor(c(
    rep("bean", length(bean_75)),
    rep("cucumber", length(cucumber_75)),
    rep("bean", length(bean_73)),
    rep("cucumber", length(cucumber_73))
  )), 
  root_extract = factor(c(
    rep("75", length(bean_75)),
    rep("75", length(cucumber_75)),
    rep("73", length(bean_73)),
    rep("73", length(cucumber_73))
  )),
  germination = c(bean_75, cucumber_75, bean_73, cucumber_73),
  batch = batch_count
)
```
(a)
```{r 2a}
model_b = glm(cbind(germination, batch - germination) ~ seed + root_extract, data = data2, family = binomial())
coef2 = summary(model_b)$coefficient
model_b |> 
  broom::tidy() |>
  knitr::kable(digits = 3)
```
**Interception: **Denoting the germination rate as $\pi$, then the model is $\log  \frac{\pi}{1-\pi}$ = `r coef2[1]` + `r coef2[2]` * is_cucumber + `r coef2[3]` * is_75media. The cucumber seeds have `r coef2[2]` units log odds ratio to successfully germinate than bean seed, while seeds cultivated with root extract media O. aegyptiaca 75 have `r coef2[3]` units log odds ratio to successfully germinate than that cultivated with O. aegyptiaca 73.

(b)
```{r 2b}
n = dim(data2)[1]
p = 3
G.stat = sum(residuals(model_b,type='pearson')^2) # pearson chisq 
phi = G.stat/(n-p)
coef3 = summary(model_b, dispersion = phi)$coefficient
model_b |> summary(dispersion = phi) 
```

```{r 2b-plot}
abs_res = abs(residuals(model_b,type='pearson')) |> sort()
qx = qnorm((n+1:n+0.5)/(2*n+1.125))
ggplot(aes(x = qx, y = res), data = tibble(
  qx = qx, res = abs_res
)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) + 
  geom_abline(intercept = 0, slope = sqrt(phi), linetype = 2, color = 'red') +
  labs(
    title = 'Half Normal Plot',
    x = 'Expected Half-Normal Order Stats',
    y ='Ordered Abs Pearson Residuals'
  ) + 
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "none"
  ) 

```
From the over-dispersion analysis and plot result, we could find that there indeed exists over-dispersion. The estimation of over-dispersion parameter is `r phi`.

**Interpretation: ** the model corrected with dispersion parameter is $\log \frac{\pi}{1-\pi}$ = `r coef2[1]` + `r coef2[2]` * is_cucumber + `r coef2[3]` * is_75media. The cucumber seeds have `r coef3[3]` units log odds ratio to successfully germinate than bean seed, while seeds cultivated with root extract media O. aegyptiaca 75 have `r coef3[3]` units log odds ratio to successfully germinate than that cultivated with O. aegyptiaca 73.

(c) Plausible cause of over-dispersion:

  1) Intra-class correlation: entire sample doesn't follow binomial distribution.

  2) Hierarchical sampling: each cluster has its own event rate which follows a distribution.

