---
title: "P8131_HW9"
author: "Pei Tian, pt2632"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F)
```

```{r lib-import}
library(tidyverse)
library(KMsurv)
library(survival)
```


# Problem 1
\(
h(x) = \frac{2x}{1+x^2}\\
H(t) = \int_0^t h(x)dx = \int_0^t \frac{1}{x^2+1}d(x^2+1) = \log (x^2+1) |_0^t = log(t^2+1)\\
S(t) = \exp\{-H(t)\} = \frac{1}{t^2+1}\\
f(t) = \frac{dS}{dt} = \frac{2t}{(t^2+1)^2}
\)

Survival function: \(S(t) = \frac{1}{t^2+1}\)
Density function: \(f(t) = \frac{2t}{(t^2+1)^2}\)

# Problem 2
Conditional probability: 
\(\hat{\lambda}_i = \text{Pr}(T = t_i | T \ge t_i) = \frac{d_i}{n_i}\)

Kaplan-Meier estimator: 
\(\hat{S}(t) = \prod_{i=1}^k (1-\hat{\lambda}_i)\)

Nelson-Aalen estimator:
\(\tilde{H}(t) = 
\begin{cases} 
0, & 0 \leq t < t_1 \\
\sum\limits_{t_i \leq t} \frac{d_i}{n_i}, & t \geq t_1 
\end{cases}
\)

Fleming-Harrington estimator: 
\(
\tilde{S}(t) = \exp\{-\tilde{H}(t)\}\)
\)

```{r data}
surv_table = tibble(
  time = c(1, 2, 4, 5, 6, 7, 8, 9, 10),
  n = c(10, 9, 7, 6, 5, 4, 3, 2, 1),
  d = c(1, 2, 0, 0, 1, 0, 0, 0, 0),
  c = c(0, 0, 1, 1, 0, 1, 1, 1, 1)
) |>
  mutate(
    lambda = d / n, 
    km_surv = cumprod(1-lambda),
    hazard = cumsum(lambda), 
    fh_surv = exp(-hazard)
  )
surv_table |> knitr::kable(digits = 2)
```

`lambda`: The hazard rate at each time point. This represents the rate at which the event is happening at each time point among the subjects at risk.

`km_surv`: The Kaplan-Meier estimate of the survival function at each time point. This gives the probability of surviving past a certain time point, given that the subject has survived up to that time point.

`hazard`: The cumulative hazard function, which estimated by Nelson-Aalen method.

`fh_surv`: This could represent the Fleming-Harrington survival estimate, which is another method to estimate the survival function, potentially giving more weight to events happening at different times.


# Problem 3

```{r surv}
data(tongue)
tongue = tongue |>
  mutate(censor = delta)

# aneuploid
fit_aneuploid = survfit(Surv(time, censor) ~ 1, data = subset(tongue, type == 1), conf.type='log')
summary(fit_aneuploid)

plot(fit_aneuploid, conf.int = T, mark.time = T, 
     xlab="Weeks", ylab="Survival Probability", 
     main="Aneuploid K-M curve", cex.lab=1.5, cex.main=1.5)

# obtain survival rate at given time, with CI
summary(fit_aneuploid, time = c(52))

# diploid
fit_diploid = survfit(Surv(time, censor) ~ 1, data = subset(tongue, type == 2), conf.type='log')
summary(fit_diploid)

plot(fit_diploid, conf.int = T, mark.time = T, 
     xlab="Weeks", ylab="Survival Probability", 
     main="Diploid K-M curve", cex.lab=1.5, cex.main=1.5)

# obtain survival rate at given time, with CI
summary(fit_diploid, time = c(52))
```
KM Curve: 
On the y-axis, we have the survival probability, which is the chance that a subject in the study will survive past a certain time. The x-axis indicates time in weeks. The curve steps down at points where events (such as death or failure) occurred. Each step downward indicates a decrease in the survival probability due to an event.

The vertical marks represent censored observations, indicating participants who were lost to follow-up or for whom the event of interest has not occurred by the end of the study period.

The dashed lines around the solid survival curve denote the 95% confidence interval, suggesting the range within which the true survival curve is likely to lie.

Comments: 

- The estimated 1-year survival probability of aneuploid is 0.654, and its 95% CI is (0.537, 0.797).

- The estimated 1-year survival probability of diploid is 0.486, and its 95% CI is (0.33, 0.716).
