---
title: "P8131_HW7"
author: "Pei Tian, pt2632"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F)
```

```{r lib-import}
library(tidyverse)
library(nlme)
library(lme4)
```

### (a)
```{r data}
data = read.csv("data/HW7-politeness_data.csv") |>
  janitor::clean_names() |>
  mutate(gender = gender |> factor(),
         scenario = scenario |> factor(),
         attitude = attitude |> factor())
```

```{r EDA1}
data |> 
  ggplot(aes(y = frequency, fill = gender)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of pitch for different gender"
  ) + 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5)) 
```
The pitch of female is significantly greater than that of male.

```{r EDA2}
data |> 
  ggplot(aes(y = frequency, fill = attitude)) +
  geom_boxplot() +
  labs(
    title = "Boxplot of pitch for different attitude"
  ) + 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))
```
There lies slight difference of pitch between pitch of population with different attitude. Population with 'inf' attitude tends to have higher pitch.



### (b)

```{r model1}
LMM1 = lme(frequency ~ gender + attitude, random = ~1 | subject,  data = data, method='REML')
summary(LMM1)
# covariance estimates for random effects
VarCorr(LMM1) 
# covariance for fixed effects 
vcov(LMM1) 
# BLUP
random.effects(LMM1) 
# residuals
data$frequency-fitted(LMM1) 
```
Given the above result, we could calculate the elements in the covariance matrix for subject \(Y_i\): 

Var(\(Y_{ij}\)) = Var(\(b_i + \epsilon_{ij}\)) = \(\sigma_{b}^2 + \sigma^2 = 598.1953 + 847.7049 = 1445.9\)

Cov(\(Y_{ij}, Y_{ik}\)) = Cov(\(b_i + \epsilon_{ij}, b_i + \epsilon_{ik}\)) = Cov(\(b_i, b_i\)) = \(\sigma_b^2 = 598.1953\)

So the covariance matrix is:
$$
\text{Cov}(Y_i) = \begin{pmatrix}
\sigma_b^2+\sigma^2 & \sigma_b^2 & \cdots & \sigma_b^2\\
\sigma_b^2 & \sigma_b^2+\sigma^2 & \cdots & \sigma_b^2\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_b^2 & \cdots & \sigma_b^2 & \sigma_b^2+\sigma^2\\
\end{pmatrix} = \begin{pmatrix}
1445.9 & 598.1953 & \cdots & 598.1953\\
598.1953 & 1445.9 & \cdots & 598.1953\\
\vdots & \vdots & \ddots & \vdots\\
598.1953 & \cdots & 598.19532 & 1445.9
\end{pmatrix}
$$
### (c)
```{r model2}
LMM2_1 = lme(frequency ~ gender + attitude, random = ~1 | subject, data = data, method='ML')
summary(LMM2_1)
LMM2_2 = lme(frequency ~ gender + attitude + gender * attitude, random = ~1 | subject, data = data, method='ML')
summary(LMM2_2)
anova(LMM2_1, LMM2_2)
```
Given p-value = 0.2392 > 0.05, the null hypothesis couldn't be rejected, which means the interaction term isn't significantly associated with the pitch in 0.05 significant level.

### (d)
```{r model3}
LMM3 = lmer(frequency ~ gender + attitude + (1 | subject) + (1 | scenario),  data = data)
summary(LMM3)
# covariance matrix for subject Y_i
VarCorr(LMM3)
```
Given the above result, we could calculate the elements in the covariance matrix for subject \(Y_i\): 

\(\sigma^2_a = 14.983^2 = 224.4903\)

\(\sigma^2_b = 24.783^2 = 614.1971\)

\(\sigma^2 = 25.254^2 = 637.7645\)

Var(\(Y_{ij}\)) = \(\sigma_{b}^2 + \sigma_{a}^2 + \sigma^2 = 1476.452\)

Cov(\(Y_{ij}, Y_{ik}\)) =  \(\sigma_b^2 + \sigma_a^2 = 838.6874\)

So the covariance matrix is:
$$
\text{Cov}(Y_i) = \begin{pmatrix}
\sigma_{a}^2 + \sigma_b^2+\sigma^2 & \sigma_{a}^2 + \sigma_b^2 & \cdots & \sigma_{a}^2 + \sigma_b^2\\
\sigma_{a}^2 + \sigma_b^2 & \sigma_{a}^2 + \sigma_b^2+\sigma^2 & \cdots & \sigma_{a}^2 + \sigma_b^2\\
\vdots & \vdots & \ddots & \vdots\\
\sigma_{a}^2 + \sigma_b^2 & \cdots & \sigma_{a}^2 + \sigma_b^2 & \sigma_{a}^2 + \sigma_b^2+\sigma^2\\
\end{pmatrix} = \begin{pmatrix}
1476.452 & 838.6874 & \cdots & 838.6874\\
838.6874 & 1476.452 & \cdots & 838.6874\\
\vdots & \vdots & \ddots & \vdots\\
838.6874 & \cdots & 838.68742 & 1476.452
\end{pmatrix}
$$

The coefficient for the fixed effect term attitude represents the expected change in the response variable for a one-unit change in attitude, holding all other variables constant. 